\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}

### 2014-07-10 paper/subdocuments/data.Rnw
minor spelling corrections

Often inequality is assessed on national level, which implies, that studies try to cover the whole population of the country of interest. This is a special thorny task for surveys working with samples, because nonresponse is a major source of bias \citep{bethlehem_handbook_2011}.  \citep{korinek_survey_2006} show, that the position in the income distribution influence the probability to participate in a survey. Low income and high income households are more likely to refuse survey response, which leads to an overrepresentation of middle income households. This process can be referred to as the "middleclass bias" \citep{diekmann_empirische_2009}. Missing data in household surveys is therefore not missing at random, which has an impact on the measures of inequality. The magnitude of this bias in Switzerland, however, is unknown. Strategies to handle this kind of bias are discussed in the literature \citep{sarndal_model_2003}, but require a register for every unit, that is proportional to income. Currently no such register exists for Switzerland \citet[43]{muller_vermogenslage_2014}. Currently used micro datasets, which are used for official publications concerning inequality in Switzerland (SILC and HABE) are furthermore confronted with a constructed coverage problem, because these surveys rely on the phone register, which excludes households not having a registered connection. \\
%(Problems of collective households, ecoplan 2014:45). Erw?hnen?

The issue of incomplete coverage is less dramatic with tax data. Essentially every permanent resident in Switzerland over 18 years of age (respectively 20 years of age prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). Basically this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover (not only) foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. Most important this includes high net wealth individuals taxed according to their expenditures. Special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below the threshold (e.g. CHF 16.900 for singles in 2010) and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units is reported, but not for the years before. Dell et al. (2007) try to estimate the fraction of non-taxed by comparing the reported numbers of tax units to census reports about the number of adult population. According to their estimations this fraction drops from 94\% in 1993/1994 to 63\% back in 1945/46. \\

>>>>

Often inequality is assessed on national level, which implies, that studies try to cover the whole population of the country of interest. This is a special thorny task for surveys working with samples, because nonresponse is a major source of bias \citep{bethlehem_handbook_2011}.  \citet{korinek_survey_2006} show, that the position in the income distribution influence the probability to participate in a survey. Low income and high income households are more likely to refuse survey response, which leads to an overrepresentation of middle income households. This process can be referred to as the "middleclass bias" \citep{diekmann_empirische_2009}. Missing data in household surveys is therefore not missing at random, which has an impact on the measures of inequality. The magnitude of this bias in Switzerland, however, is unknown. Strategies to handle this kind of bias are discussed in the literature \citep{sarndal_model_2003}, but require a register for every unit, that is proportional to income. Currently no such register exists for Switzerland \citet[43]{muller_vermogenslage_2014}. Currently used micro datasets, which are used for official publications concerning inequality in Switzerland (SILC and HABE) are furthermore confronted with a constructed coverage problem, because these surveys rely on the phone register, which excludes households not having a registered connection. \\
%(Problems of collective households, ecoplan 2014:45). Erw?hnen?

The issue of incomplete coverage is less dramatic with tax data. Essentially every permanent resident in Switzerland over 18 years of age (respectively 20 years of age prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). Basically this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover (not only) foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. Most important this includes high net wealth individuals taxed according to their expenditures. Special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below the threshold (e.g. CHF 16.900 for singles in 2010) and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units is reported, but not for the years before. citet{dell_income_2005} try to estimate the fraction of non-taxed by comparing the reported numbers of tax units to census reports about the number of adult population. According to their estimations this fraction drops from 94\% in 1993/1994 to 63\% back in 1945/46. \\

### 2014-07-10 paper/subdocuments/methods.Rnw
minor spelling corrections

To fulfill the above described task, we use two techniques. To assess the development over time, we calculate in general Gini-coefficients for all possible time points, allowing us the make time trends visible and to compare the result out of the FTA-tax data to other publications. For selected periods we expand the analysis with relative distribution methods, which allows an in-depth analysis of distributional differences and therefore catches the shortcomings of Gini-coefficients. \\

@@

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe how the two distributions are gone be transformed into a relative distribution, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $x$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(x)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $x$. This function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This already allows to spot distributional differences over the whole scale of $x$ visually. \\

@@

% R based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index developed by \citep{Handock and Morris 1999} addresses this issue, because it is decomposable to distinguish differences in the upper and lower tails.  Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

>>>>

To fulfill the above described task, we use two techniques. To assess the development over time, we calculate in general Gini-coefficients for all possible time points, allowing us to make time trends visible and to compare the result out of the FTA-tax data to other publications. For selected periods we expand the analysis with relative distribution methods, which allows an in-depth analysis of distributional differences and therefore catches the shortcomings of Gini-coefficients. \\

@@

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe how the two distributions are going to be transformed into a relative distribution, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $x$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(x)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $x$. This function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This already allows to spot distributional differences over the whole scale of $x$ visually. \\

@@

% R based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index developed by \citep{handcock_relative_1999} addresses this issue, because it is decomposable to distinguish differences in the upper and lower tails.  Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\
